---
title: "class10"
author: "Ziheng Huang"
format: html
---

# 1. Importing candy data

```{r}
candy_file <- 'https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv'
candy = read.csv(candy_file, row.names=1)
head(candy)
```

```{r}
dim(candy)
```

Q1. How many different candy types are in this dataset? 12 categories, 85 brands of candy

Q2. How many fruity candy types are in the dataset? 38

```{r}
sum(candy$fruity)
```

# 2. What is your favorate candy?

Q3. What is your favorite candy in the dataset and what is it's winpercent value? Twizzlers:

    45.46628

```{r}
candy["Twizzlers", ]$winpercent
```

Q4. What is the winpercent value for "Kit Kat"?

    76.7686

```{r}
candy["Kit Kat", ]$winpercent
```

Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

    49.6535

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

Q9. Is the distribution of winpercent values symmetrical?

It is roughly symmetrical a bit right skewed.

Q10. Is the center of the distribution above or below 50%?

Center is below 50.

Q11. On average is chocolate candy higher or lower ranked than fruit candy? chocolate is higher

```{r}
# chocolate rank mean
mean(candy$winpercent[as.logical(candy$chocolate)])
# fruit rank mean
mean(candy$winpercent[as.logical(candy$fruit)])
```

Q12. Is this difference statistically significant? it is statistically significant, p \< 0.05

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],candy$winpercent[as.logical(candy$fruit)])
```

# 3. Overall Candy Rankings

Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```

Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
aes(winpercent, reorder(rownames(candy),winpercent)) +
geom_col()
```

### Time to add some useful color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Q17. What is the worst ranked chocolate candy? Charleston Chew

Q18. What is the best ranked fruity candy? Nik L Nip

# 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Tootsie Roll Midgies have the highest winpercent/price ratio.

```{r}
ord <- order(candy$winpercent/candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip is the least popular

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called "dot chat" or "lollipop" chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

# 5 Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity & chocolate, caramel & fruity, peanutyalmondy & fruity, nougat & fruity, crispedricewafer & fruity, bar & fruity, pricepercent & fruity, winpercent & fruity, hard & caramel, hard & peanuty, hard & nougat, hard & crispedricewafer, crispedricewafer & nougat, bar & hard, etc... all the pairs where the color is red.

Q23. Similarly, what two variables are most positively correlated?

bar & chocolate, winpercent & chocolate, pricepercent & bar

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot( pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2")
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)

```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, pluribus are picked up. They make sense.
